### 2015-03-13 doc/notebook/2015-03-13_Archaeal_data_acquisition.Rmd
Adapt for archaea

# Getting the bacterial sequence data and metadata

@@

@@

To get good sequences...
* Go to search window
* Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
* Hit Mark Listed, Unmark Rest button (N=1493493)
* The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure
    out which taxonomy to base the analysis on. Do the following searches...
    * ARB_color != 1 & tax_rdp == "*Bacteria*" (N=1324558)
    * ARB_color != 1 & tax_greengenes == "*Bacteria*" (N=1123144)
    * ARB_color != 1 & tax_slv == "*Bacteria*" (N=426917)
    * ARB_color != 1 & tax_embl == "*Bacteria*" (N=1207210)
    * tax_rdp | tax_greengenes | tax_embl | tax_slv (N=1376910)

Let's stick with the bacterial sequences that have an RDP taxonomy and we will
analyze them in RDP space, but preserve the taxonomy generated by the other
systems. Nicely, the RDP taxonomy does not include mitochondrial
16S rRNA gene sequences, so all we'll need to remove are the chloroplast
sequences

* Return to Search and Query window
* Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
* Change the second search field to "tax_rdp" and set it to "*Bacteria*"
* Change the third search field to "tax_rdp" and set it to "*Chloroplast*"
* Set the two Query operators to "and" and hit "Search"
* Hit Mark Listed Unmark Rest (N=1,301,541)
* Click "Write to Fields of Listed", select the "remark" field, and enter "good bacteria"
* Click "Write"
* In main ARB window go File->Export->Export to external format
* Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
* Rename `noname.fasta` to `bacteria.fasta`

Now we need the taxonomy information.
* Go Tree -> NDS
* Click "name", "acc", "tax_rdp". The "tax_rdp" field should have 250 characters
* Unclick everything else
* Click "Close"
* Go File->Export->Export fields
* Set the file name to "bacteria.taxonomy" and Column output to "TAB separated"
* Click "SAVE"

@@

The next thing we need to do is to clean up the bacteria.taxonomy.nds file
to make it into a proper, mothur compatible bacteria.taxonomy file. First we
want to remove the names of any sub taxa (e.g. suborder). To do this we need
to get the RDP list of names and the taxonomic level they belong to...

@@

```{r format_taxonomy}
tax_data <- read.table(file="bacteria.taxonomy.nds", sep="\t")

@@

#function to see which taxa names in a list are found in the sub taxa list
remove.subs <- function(tax.vector){
,	#return(tax.vector[which(!tax.vector %in% sub.names)])
}

@@

write.table(cbind(seq_names, no.subs.str), "bacteria.tax", row.names=F,
                                                col.names=F, quote=F, sep="\t")
```

@@

```{r get_summary, engine="bash", results="hide", echo=TRUE}
mothur "#summary.seqs(fasta=bacteria.fasta, processors=12)"
```

```
                Start   End     NBases  Ambigs  Polymer NumSeqs
Minimum:        1       9693    900     0       4       1
2.5%-tile:      1004    12895   1279    0       5       32539
25%-tile:       1026    13069   1355    0       5       325386
Median:         1044    13267   1382    0       5       650771
75%-tile:       1051    13825   1447    0       6       976156
97.5%-tile:     1793    13991   1522    2       7       1269003
Maximum:        6448    15767   3281    51      38      1301541
Mean:   1096.18 13442   1399.39 0.176343        5.49738
# of Seqs:      1301541
```

Notes...
* Looks like an archaeal sequence slipped through
* Will want to get rid of sequences with large number of Ns in them - how large?

```{r find_screening_points}
data <- read.table(file="bacteria.summary", header=T, row.names=1)
quantile(data$start, probs=seq(0.9,1,0.01))
start <- 2000

quantile(data$end, probs=seq(0,0.1,0.01))
end <- 13000

quantile(data$nbases, probs=seq(0.0,0.1,0.01))
min_length <- 1200

@@

trimmed <- data[data$start <= start & data$end >= end &
                    data$nbases > min_length & data$ambigs <= max_ambig,]
nrow(trimmed)
summary(trimmed$nbases)
```

@@

```{r screen_data, engine="bash"}
mothur "#set.current(processors=8);
        set.dir(input=data/raw, output=data/process)
        screen.seqs(fasta=bacteria.fasta, taxonomy=bacteria.tax, start=2000, end=13000, maxambig=2, minlength=1200);"
```

@@

```{r single_cell, engine="bash"}
mothur "#align.seqs(fasta=single_cell.bacteria.fasta, reference=bacteria.good.fasta);
        screen.seqs(fasta=current, start=2000, end=13000, maxambig=2, minlength=1200);
        classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
```

@@

```{r pool_data, engine="bash"}
cat bacteria.good.fasta single_cell.bacteria.good.align > all_bacteria.align
cat bacteria.good.tax single_cell.bacteria.good.pds.wang.taxonomy > all_bacteria.taxonomy

mothur "#filter.seqs(fasta=all_bacteria.align, vertical=T, trump=.);
        unique.seqs();
        pre.cluster(fasta=current, name=current, diffs=10);"

#need to make the taxonomy file match the fasta and names files...
cut -f 1 all_bacteria.filter.unique.precluster.names > precluster.accnos
mothur "#get.seqs(taxonomy=all_bacteria.taxonomy, accnos=precluster.accnos)"
```

@@

```{r cluster_data, engine="bash"}
mothur "#cluster.split(fasta=all_bacteria.filter.unique.precluster.fasta, name=all_bacteria.filter.unique.precluster.names, taxonomy=all_bacteria.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
```

@@

We'd like to use some metadata from the database to characterize the changes in
the representation of sequences over time, by environment, methods, etc. The
fields housed within SILVA are available at their [website](http://www.arb-silva.de/fileadmin/arb_web_db/release_115/Fields_description/SILVA_description_of_fields_16_06_2013.htm)
and there is an [FAQ](http://www.arb-silva.de/documentation/faqs/) on their site
as well. There were a number of fields that I didn't think were relevant and
instead focused on 30 fields that I thought could help the cause. I marked these
fields in the NDS feature and extracted them to `metadata.nds`. We need to
tweak this file slightly in R to get it into a format that we can use.

@@

Now we want to read in `bacteria.bad.accnos` so that we can figure out which
sequences to cull from the metadata table.

```{r}
bad_accnos_data <- read.table(file="data/process/bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
bad_accnos <- bad_accnos_data$name.accnos
metadata_good <- metadata[-c(bad_accnos),]
write.table(metadata_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t")
```

We'd also like to get the metadata from the single cell data. We have an `xlsx` that the LSU group pulled together that we'll read in and concatenate to the end of `bacteria.good.metadata`.

@@

spread_sheet <- read.xlsx("data/raw/BacterialSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
bad_accnos_data <- read.table(file="data/process/single_cell.bacteria.bad.accnos", col.names=c("name.accnos", "reason"))
bad_accnos <- bad_accnos_data$name.accnos
spread_sheet_good <- spread_sheet[-c(bad_accnos),]

write.table(spread_sheet_good, file="data/process/all_bacteria.good.metadata", quote=F, sep="\t", col.names=F, append=T)
```

>>>>

# Getting the archaeal sequence data and metadata

@@

This is redundant with what we saw for the bacterial data, but for completion...

@@

To get good sequences...
* Go to search window
* Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
* Hit Mark Listed, Unmark Rest button (N=1493493)
* The problem with the taxonomies is that the sequences don't all have taxonomies. Need to figure
    out which taxonomy to base the analysis on. Do the following searches...
    * ARB_color != 1 & tax_rdp == "*Archaea*" (N=40288)
    * ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19795)
    * ARB_color != 1 & tax_slv == "*Archaea*" (N=17641)
    * ARB_color != 1 & tax_embl == "*Archaea*" (N=44574)
    * ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=46387 )

Let's stick with the archaeal sequences that have an RDP taxonomy and we will
analyze them in RDP space, but we can always go back to the taxonomy provided by
the other systems if we'd like. A quick spot checking suggests that the RDP
taxonomy information is actually richer than that of the EBML taxonomy
information.

* Return to Search and Query window
* Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
* Change the second search field to "tax_rdp" and set it to "*Archaea*"
* Hit "Search"
* Hit Mark Listed Unmark Rest (N=40,288)
* Click "Write to Fields of Listed", select the "remark" field, and enter "good archaea"
* Click "Write"
* In main ARB window go File->Export->Export to external format
* Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
* Rename `noname.fasta` to `archaea.fasta`

Now we need the taxonomy information.
* Go Tree -> NDS
* Click "name", "acc", "tax_rdp". The "tax_rdp" field should have 250 characters
* Unclick everything else
* Click "Close"
* Go File->Export->Export fields
* Set the file name to "archaea.taxonomy" and Column output to "TAB separated"
* Click "SAVE"

@@

The next thing we need to do is to clean up the archaea.taxonomy.nds file
to make it into a proper, mothur compatible archaea.taxonomy file. First we
want to remove the names of any sub taxa (e.g. suborder). To do this we need
to get the RDP list of names and the taxonomic level they belong to...

@@

```{r format_taxonomy}
tax_data <- read.table(file="archaea.taxonomy.nds", sep="\t")

@@

#function to see which taxa names in a list are found in the sub taxa list
remove.subs <- function(tax.vector){
    return(tax.vector[which(!tax.vector %in% sub.names)])
}

@@

write.table(cbind(seq_names, no.subs.str), "archaea.tax", row.names=F,
                                                col.names=F, quote=F, sep="\t")
```

@@

```{r get_summary, engine="bash", results="hide", echo=TRUE}
mothur "#summary.seqs(fasta=archaea.fasta, processors=12)"
```

```
            Start    End	NBases	Ambigs	Polymer	NumSeqs
Minimum:	1        4058	900	    0    	4	1
2.5%-tile:	357      4692	905	    0    	5	1008
25%-tile:	365      4783	935	    0   	5	10073
Median: 	379      7409	1262	0   	5	20145
75%-tile:	456      7729	1349	0    	6	30217
97.5%-tile:	1746	 7984	1470	4   	7	39281
Maximum:	2321	 8711	2174	46     	32	40288
Mean:	    513.659	 6698.2	1185.3	0.3	    5.44244
# of Seqs:	40288
```

Notes...
* Will want to get rid of sequences with large number of Ns in them and the
ridiculously long homopolymrs - how large?

```{r find_screening_points}
data <- read.table(file="archaea.summary", header=T, row.names=1)
quantile(data$start, probs=seq(0.9,1,0.01))
start <- 2000

quantile(data$end, probs=seq(0,0.1,0.01))
end <- 4058

quantile(data$nbases, probs=seq(0.0,0.1,0.01))
min_length <- 900

@@

quantile(data$polymer, probs=seq(0.9,1.0,0.01))
max_polymer <- 8

trimmed <- data[data$start <= start & data$end >= end &
                    data$polymer <= max_polymer & data$nbases > min_length &
                    data$ambigs <= max_ambig,]
nrow(trimmed)
summary(trimmed$nbases)
```

@@

```{r screen_data, engine="bash"}
mothur "#set.current(processors=8);
        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);"
```

@@

```{r single_cell, engine="bash"}
mothur "#align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
        screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
        classify.seqs(fasta=current, reference=trainset10_082014.pds.fasta, taxonomy=trainset10_082014.pds.tax, cutoff=80, processors=8)"
```

@@

```{r pool_data, engine="bash"}
cat archaea.good.fasta single_cell.archaea.good.align > all_archaea.align
cat archaea.good.tax single_cell.archaea.good.pds.wang.taxonomy > all_archaea.taxonomy

mothur "#filter.seqs(fasta=all_archaea.align, vertical=T, trump=.);
        unique.seqs();
        pre.cluster(fasta=current, name=current, diffs=9);"

#need to make the taxonomy file match the fasta and names files...
cut -f 1 all_archaea.filter.unique.precluster.names > precluster.accnos
mothur "#get.seqs(taxonomy=all_archaea.taxonomy, accnos=precluster.accnos)"
```

@@

```{r cluster_data, engine="bash"}
mothur "#cluster.split(fasta=all_archaea.filter.unique.precluster.fasta, name=all_archaea.filter.unique.precluster.names, taxonomy=all_archaea.pick.taxonomy, taxlevel=2, classic=T, processors=8)"
```

@@

We'd like to use some metadata from the database to characterize the changes in
the representation of sequences over time, by environment, methods, etc. The
fields housed within SILVA are available at their [website](http://www.arb-silva.de/fileadmin/arb_web_db/release_115/Fields_description/SILVA_description_of_fields_16_06_2013.htm)
and there is an [FAQ](http://www.arb-silva.de/documentation/faqs/) on their site
as well. There were a number of fields that I didn't think were relevant and
instead focused on 30 fields that I thought could help the cause. I marked these
fields in the NDS feature and extracted them to `archaea_metadata.nds`. We need to
tweak this file slightly in R to get it into a format that we can use.

@@

Now we want to read in `archaea.bad.accnos` so that we can figure out which
sequences to cull from the metadata table.

```{r}
bad_accnos_data <- read.table(file="data/process/archaea.bad.accnos", col.names=c("name.accnos", "reason"))
bad_accnos <- bad_accnos_data$name.accnos
metadata_good <- metadata[-c(bad_accnos),]
write.table(metadata_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t")
```

We'd also like to get the metadata from the single cell data. We have an `xlsx` that the LSU group pulled together that we'll read in and concatenate to the end of `archaea.good.metadata`.

@@

spread_sheet <- read.xlsx("data/raw/ArchaealSSUMetadata.xlsx", sheet=1, startRow=1, colNames=TRUE)
bad_accnos_data <- read.table(file="data/process/single_cell.archaea.bad.accnos", col.names=c("name.accnos", "reason"))
bad_accnos <- bad_accnos_data$name.accnos
spread_sheet_good <- spread_sheet[-c(bad_accnos),]

write.table(spread_sheet_good, file="data/process/all_archaea.good.metadata", quote=F, sep="\t", col.names=F, append=T)
```
